\large{\bf Looking forwards and backwards through locally regulated populations }

### 2023-03-22 forwards-and-backwards.tex
D notation

	\comment{Need to thank KAVLI, Paris Cit\'e, various funders}
    \comment{make notation for $\mathcal{D}$ consistent}

@@

\begin{definition}[Martingale Problem Characterisation]
    \label{defn:lookdown_mgale}
For given positive values of $N$ and $\theta$, define the generator $A^N$
by
\begin{equation}
	\label{eqn:lookdown_generator}
\begin{split}
& A^{N}f(\lp ) \\
&\quad =
    f(\lp)
    \,\sum_{(x,u)\in \lp}\,
    2 \gamma(x, \eta)
    \Bigg\{ \frac{1}{2N} \int_u^N g(x,u_1) du_1
            \frac{
                \theta \int_{\IR^d}
                (g(y,u) - g(x,u))
                r(y, \eta) q_{\theta}(x,dy)
            }{ g(x,u) }
        \\
    &\qquad\qquad\qquad\qquad\qquad\qquad\qquad {} +
        \frac{\theta}{N} \int_u^N
        \int_{\IR^d}\left(
            \frac{ g(y,u_1) + g(x,u_1) }{ 2 } - 1
        \right)
        r(y, \eta) q_{\theta}(x,dy)
        du_1
    \Bigg\}\\
    &\qquad\qquad {} +
    f(\lp) \sum_{(x,u)\in\lp}\,
    \left(
        c_\theta(x,\eta) u^2 - b_{\theta}(x,\eta)u
    \right)
    \frac{\partial_u g(x,u)}{g(x,u)}
    ,
\end{split}
\end{equation}
where $f(\lp) = \prod_{(x, u) \in \lp} g(x, u)$ is as defined in \eqref{eqn:test_functions},
and $\eta(\cdot) = \lp(\cdot \times [0, N]) / N$
as before. 
Given $\lp_0 \in {\mathcal C}_N$, 
we say that a ${\mathcal D}([0,\infty), {\mathcal C}_N)$-valued process 
$(\xi_t)_{t \ge 0}$
   is a solution to the $(A^N, \lp_0)$ martingale problem
    if 
$f(\lp_t) - f(\lp_0)-\int_0^t A^N f(\lp_s) ds$ is a martingale (with 
respect to the natural filtration) for all 
test functions $f$ 
    as defined above.
\end{definition}

@@

\begin{proof}[Proof of Theorem~\ref{thm:nonlocal_convergence}.]
The proof follows a familiar pattern. First we extend $\IR^d$ 
to its one-point compactification $\overline{\IR}^d$ and 
establish, in Lemma~\ref{lem:eta_compact_containment},
compact containment of the 
sequence of scaled population
processes in $\cmeasures$ 
(for which, since we have compactified $\IR^d$,
it suffices to consider the sequence of total masses); armed with this, 
%Theorem~3.9.1 of~\cite{ethier/kurtz:1986} tells us that 
tightness of the population processes in $\mathcal{D}_{\cmeasures}([0,\infty))$
follows from tightness of the real-valued processes $(H(\eta_t))_{t\geq 0}$ for a sufficiently large
class of test functions $H$, which we establish through an application of the
Aldous-Rebolledo criterion in Lemma~\ref{lem:eta_projections_tightness}.
These ingredients are gathered together in 
Proposition~\ref{tightness in one point compactification}
to deduce tightness of the scaled population processes in 
the larger space $\mathcal{D}_{\cmeasures}([0,\infty))$.

We then characterise
limit points as solutions to a martingale problem in
Lemma~\ref{lem:limit_mgale}; finally in Lemma~\ref{no mass at infty}
we check that 
in the process of passing to the limit, no mass `escaped to infinity', so that
in fact the limit points take values in $\mathcal{D}_{\measures}([0,\infty))$.
\end{proof}

@@

\begin{proof}
Consider the semimartingale decomposition from equation~\eqref{eqn:eta_martingale}:
\begin{align} \label{eqn:eta_f_mgale_decomp}
    \langle f, \eta^N_t \rangle
    &=
    \langle f, \eta^N_0 \rangle
    + \int_0^t \int_{\IR^d} \big\{
        \gamma(x, \eta^N_s) B_f(x, \eta^N_s)
        + f(x) F(x, \eta^N_s)
        \big\} \eta^N_s(dx) ds
    + M^N_t(f) ,
\end{align}
where $M^N_t(f)$ is a martingale.
% and we are defining
% \begin{equation}
% \label{generator Bf}
%     B_f(x, \eta) = \int_{\IR^d} \theta (f(y) r(y, \eta) - f(x) r(x, \eta)) q_\theta(x, dy) .
% \end{equation}

@@

\begin{proof}[of Proposition~\ref{thm:mmt_application}]
    This proposition is the content of the Markov Mapping Theorem,
    Theorem A.2 in \citet{etheridge/kurtz:2019},
    applied to our situation.
    \comment{TODO: copy the MMT into our appendix and refer to it}
    In the notation of that theorem, $\gamma$ is what we have called $\kappa$ above,
    $\alpha$ is the transition function that assigns levels uniformly on $[0, N]$ (in the first case)
    or as a Poisson process with Lesbegue intensity (in the limiting case),
    and we need a continuous $\psi^N(\lp)$ such that $|A^N f(\lp)| \le c_f \psi^N(\lp)$
    for all $f \in \mathcal{D}(A^N)$ (and similarly a function $\psi$ for $A$).
    We also need that applying the lookdown generator to a function and averaging over levels
    is equivalent to applying the population process generator to the function
    whose dependence on levels has been averaged out,
    a condition which we precisely state, and verify,
    in Appendix~\ref{sec: Markov Mapping Theorem Application}.

@@

\begin{theorem}
\label{teo: lookdown on pre-limit well defined}
For each $N \in \mathbb{N}$,
there exists a solution $(\lp^N_t)_{t \geq 0}$ to the $(A^N, \lp^N_0)$ martingale problem.
Furthermore, if we define $\Xi^N_t = \frac{1}{N}\lp^N_t(dx \times [0,N])$ for $t \geq 0$,
then the pre-limiting model $(\eta^N_t(dx))_{t \geq 0}$ defined in Definition \ref{defn:mgale_construction}
and $(\Xi^N_t(dx))_{t \geq 0}$
shares the same distribution on $\mathcal{D}_{\mathcal{M}(\mathbb{R}^d)}[0,\infty)$.

@@

Furthermore, if we define $\Xi_t = \lim_{u_0 \to \infty} \frac{1}{u_0}\lp_t(dx \times [0,u_0])$,
then the limit model $(\eta_t(dx))_{t \geq 0}$ defined in Theorem \ref{thm:nonlocal_convergence}
and $(\Xi_t(dx))_{t \geq 0}$
shares the same distribution on $\mathcal{D}_{\mathcal{M}(\mathbb{R}^d)}[0,\infty)$.

@@

As a result,
we can apply Markov Mapping Theorem to establish the following statements:
\begin{enumerate}
    \item There exists a $\mathcal{M}_F(\mathbb{R}^d\times[0,N])$-valued process $(\lp^N_t)_{t \geq 0}$
    such that 
    $M^N_t:=    f(\xi^N_t)-   f(\xi^N_0)-\int_{0}^{t}  A^Nf(\xi^N_s)ds$ is a martingale for all $f\in D(A^N)$,
    \item The process $(\gamma_M \circ \xi^N_t)_{t \geq 0}= \left(\Xi^N_t\right)_{t \geq 0}$ 
    has the same distribution in $\mathcal{D}_{\mathcal{M}(\mathbb{R}^d)}[0,\infty)$ as $(\eta^N_t)_{t \geq 0}$ defined in Definition \ref{def: MP definition of pre-limit}.
    \item $\mathbb{P}\{\xi^N_t \in B | \mathcal{F}^{\Xi^N}_t\}=\alpha_M(\Xi^N(t),B), ~~~B \in \mathcal{B}(\mathcal{M}(\IR^d \times [0,N])).$
\end{enumerate}
Note that in particular,
from the third statement,
we can establish that for $h(x,u) \in \mathcal{C}_{c}(\mathbb{R}^d \times[0,N])$ and $h \geq 0$,

>>>>

	\comment{Need to thank KAVLI, Paris Cit\'e, various funders}

@@

\begin{definition}[Martingale Problem Characterisation]
    \label{defn:lookdown_mgale}
For given positive values of $N$ and $\theta$, define the generator $A^N$
by
\begin{equation}
	\label{eqn:lookdown_generator}
\begin{split}
& A^{N}f(\lp ) \\
&\quad =
    f(\lp)
    \,\sum_{(x,u)\in \lp}\,
    2 \gamma(x, \eta)
    \Bigg\{ \frac{1}{2N} \int_u^N g(x,u_1) du_1
            \frac{
                \theta \int_{\IR^d}
                (g(y,u) - g(x,u))
                r(y, \eta) q_{\theta}(x,dy)
            }{ g(x,u) }
        \\
    &\qquad\qquad\qquad\qquad\qquad\qquad\qquad {} +
        \frac{\theta}{N} \int_u^N
        \int_{\IR^d}\left(
            \frac{ g(y,u_1) + g(x,u_1) }{ 2 } - 1
        \right)
        r(y, \eta) q_{\theta}(x,dy)
        du_1
    \Bigg\}\\
    &\qquad\qquad {} +
    f(\lp) \sum_{(x,u)\in\lp}\,
    \left(
        c_\theta(x,\eta) u^2 - b_{\theta}(x,\eta)u
    \right)
    \frac{\partial_u g(x,u)}{g(x,u)}
    ,
\end{split}
\end{equation}
where $f(\lp) = \prod_{(x, u) \in \lp} g(x, u)$ is as defined in \eqref{eqn:test_functions},
and $\eta(\cdot) = \lp(\cdot \times [0, N]) / N$
as before. 
Given $\lp_0 \in {\mathcal C}_N$, 
we say that a ${\mathcal D}_{[0,\infty)}({\mathcal C}_N)$-valued process 
$(\xi_t)_{t \ge 0}$
   is a solution to the $(A^N, \lp_0)$ martingale problem
    if 
$f(\lp_t) - f(\lp_0)-\int_0^t A^N f(\lp_s) ds$ is a martingale (with 
respect to the natural filtration) for all 
test functions $f$ 
    as defined above.
\end{definition}

@@

\begin{proof}[Proof of Theorem~\ref{thm:nonlocal_convergence}.]
The proof follows a familiar pattern. First we extend $\IR^d$ 
to its one-point compactification $\overline{\IR}^d$ and 
establish, in Lemma~\ref{lem:eta_compact_containment},
compact containment of the 
sequence of scaled population
processes in $\cmeasures$ 
(for which, since we have compactified $\IR^d$,
it suffices to consider the sequence of total masses); armed with this, 
%Theorem~3.9.1 of~\cite{ethier/kurtz:1986} tells us that 
tightness of the population processes in $\mathcal{D}_{[0,\infty)}(\cmeasures)$
follows from tightness of the real-valued processes $(H(\eta_t))_{t\geq 0}$ for a sufficiently large
class of test functions $H$, which we establish through an application of the
Aldous-Rebolledo criterion in Lemma~\ref{lem:eta_projections_tightness}.
These ingredients are gathered together in 
Proposition~\ref{tightness in one point compactification}
to deduce tightness of the scaled population processes in 
the larger space $\mathcal{D}_{[0,\infty)}(\cmeasures)$.

We then characterise
limit points as solutions to a martingale problem in
Lemma~\ref{lem:limit_mgale}; finally in Lemma~\ref{no mass at infty}
we check that 
in the process of passing to the limit, no mass `escaped to infinity', so that
in fact the limit points take values in $\mathcal{D}_{[0,\infty)}(\measures)$.
\end{proof}

@@

\begin{proof}
Consider the semimartingale decomposition from equation~\eqref{eqn:eta_martingale}:
\begin{align} \label{eqn:eta_f_mgale_decomp}
    \langle f, \eta^N_t \rangle
    &=
    \langle f, \eta^N_0 \rangle
    + \int_0^t \int_{\IR^d} \big\{
        \gamma(x, \eta^N_s) B_f(x, \eta^N_s)
        + f(x) F(x, \eta^N_s)
        \big\} \eta^N_s(dx) ds
    + M^N_t(f) ,
\end{align}
where $M^N_t(f)$ is a martingale
and $B_f$ is defined in \comment{TODO REFER}.
% and we are defining
% \begin{equation}
% \label{generator Bf}
%     B_f(x, \eta) = \int_{\IR^d} \theta (f(y) r(y, \eta) - f(x) r(x, \eta)) q_\theta(x, dy) .
% \end{equation}

@@

\begin{proof}[of Proposition~\ref{thm:mmt_application}]
    This proposition is the content of the Markov Mapping Theorem,
    Theorem A.2 in \citet{etheridge/kurtz:2019},
    applied to our situation.
    \comment{TODO: copy the MMT into our appendix and refer to it}
    In the notation of that theorem, $\gamma$ is what we have called $\kappa$ above,
    $\alpha$ is the transition function that assigns levels uniformly on $[0, N]$ (in the first case)
    or as a Poisson process with Lesbegue intensity (in the limiting case),
    and we need a continuous $\psi^N(\lp)$ such that $|A^N f(\lp)| \le c_f \psi^N(\lp)$
    for all $f$ in the domain of $A^N$ (and similarly a function $\psi$ for $A$).
    We also need that applying the lookdown generator to a function and averaging over levels
    is equivalent to applying the population process generator to the function
    whose dependence on levels has been averaged out,
    a condition which we precisely state, and verify,
    in Appendix~\ref{sec: Markov Mapping Theorem Application}.

@@

\begin{theorem}
\label{teo: lookdown on pre-limit well defined}
For each $N \in \mathbb{N}$,
there exists a solution $(\lp^N_t)_{t \geq 0}$ to the $(A^N, \lp^N_0)$ martingale problem.
Furthermore, if we define $\Xi^N_t = \frac{1}{N}\lp^N_t(dx \times [0,N])$ for $t \geq 0$,
then the pre-limiting model $(\eta^N_t(dx))_{t \geq 0}$ defined in Definition \ref{defn:mgale_construction}
and $(\Xi^N_t(dx))_{t \geq 0}$
shares the same distribution on $\mathcal{D}_{[0,\infty)}(\mathcal{M}(\mathbb{R}^d))$.

@@

Furthermore, if we define $\Xi_t = \lim_{u_0 \to \infty} \frac{1}{u_0}\lp_t(dx \times [0,u_0])$,
then the limit model $(\eta_t(dx))_{t \geq 0}$ defined in Theorem \ref{thm:nonlocal_convergence}
and $(\Xi_t(dx))_{t \geq 0}$
shares the same distribution on $\mathcal{D}_{[0,\infty)}(\mathcal{M}(\mathbb{R}^d))$.

@@

As a result,
we can apply Markov Mapping Theorem to establish the following statements:
\begin{enumerate}
    \item There exists a $\mathcal{M}_F(\mathbb{R}^d\times[0,N])$-valued process $(\lp^N_t)_{t \geq 0}$
    such that 
    $M^N_t:=    f(\xi^N_t)-   f(\xi^N_0)-\int_{0}^{t}  A^Nf(\xi^N_s)ds$ is a martingale for all $f\in D(A^N)$,
    \item The process $(\gamma_M \circ \xi^N_t)_{t \geq 0}= \left(\Xi^N_t\right)_{t \geq 0}$ 
    has the same distribution in $\mathcal{D}_{[0,\infty)}(\mathcal{M}(\mathbb{R}^d))$ as $(\eta^N_t)_{t \geq 0}$ defined in Definition \ref{def: MP definition of pre-limit}.
    \item $\mathbb{P}\{\xi^N_t \in B | \mathcal{F}^{\Xi^N}_t\}=\alpha_M(\Xi^N(t),B), ~~~B \in \mathcal{B}(\mathcal{M}(\IR^d \times [0,N])).$
\end{enumerate}
Note that in particular,
from the third statement,
we can establish that for $h(x,u) \in \mathcal{C}_{c}(\mathbb{R}^d \times[0,N])$ and $h \geq 0$,
